# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data
```{r}
activity <- read.csv(unzip("activity.zip"))
activity$date <- as.Date(activity$date,"%Y-%m-%d")
```

## What is mean total number of steps taken per day?
```{r}
# Split the activities per day
days <- cut(activity$date,"day")
daily.activity <- split(activity,days)

# Calculate and plot the daily total steps
steps.sum <- sapply(daily.activity,function(x){sum(x$steps,na.rm=TRUE)})
hist(steps.sum,breaks=10,main="Total Daily Steps",xlab="Daily steps")

# Report mean and median total daily steps
steps.mean <- round(mean(steps.sum))
steps.median <- median(steps.sum)

print(paste("Total daily steps mean =",steps.mean,", median =",steps.median))
```


## What is the average daily activity pattern?
```{r}
# Split the data by the 5-minutes interval across all days
activity.5min <- split(activity, activity$interval)

# Calculate and plot the average steps per every 5-minutes interval
mean.5min <- sapply(activity.5min,function(x) { round(mean(x$steps,na.rm=T))})
plot(unique(activity$interval), mean.5min, type="l",main="Daily Activity Pattern",xlab="5-minutes interval",ylab="Average Steps")
max.interval <- names(mean.5min[which.max(mean.5min)])
max.value <- mean.5min[which.max(mean.5min)]
```
The interval with maximum number of steps is `r max.interval`, with average steps = `r max.value`

## Imputing missing values
``` {r}
# Get the rows with incomplete cases i.e. has NA
incomplete <- activity[!complete.cases(activity),]
na.count <- nrow(incomplete)

# Create a new data.frame from activity
activity.complete <- activity

# Get intervals with incomplete data
intervals <- unique(incomplete$interval)

# Fill missing data with the interval mean that was calculated previously i.e. mean.5min
for(i in 1:length(intervals)){
  incomplete[incomplete$interval == intervals[i],"steps"] <- mean.5min[which(names(mean.5min)==intervals[i])]
}

activity.complete[!complete.cases(activity.complete),] <- incomplete # now already completed

new.steps.sum <- sapply(split(activity.complete,days),function(x){sum(x$steps)})
hist(new.steps.sum,breaks=10,main="Total Daily Steps with Imputed Values",xlab="Daily steps")

# Report mean and median total daily steps
new.steps.mean <- round(mean(new.steps.sum))
new.steps.median <- median(new.steps.sum)

print(paste("Total daily steps mean =",new.steps.mean,", median =",new.steps.median))
```

It is clear that after imputing missing values the total number of steps increased particularly the daily total steps mean increased by **`r new.steps.mean - steps.mean`** and the median increased by **`r new.steps.median - steps.median`** steps. Also, the histogram shape is more close to the normal distribution.


## Are there differences in activity patterns between weekdays and weekends?
